Continous time Markov chain models; \\ extension, simulation and visualisation

### 2015-02-11 Response.tex

\begin{itemize}

% 1
\item {
{\it
\textbf{
The thesis starts with a very nice introduction, which I think is very insightful for the reader who is not familiar with phylogenetic and/or Bayesian inference. 
In spite of this, I believe that the readability of the later chapters would be further improved if the introduction:
}% END: textbf
}% END: it
}% END: item

\begin{itemize}

% 1a
\item {
{\it
\textbf{
Briefly described the evolutionary models, such as GY94, that are later used.
}% END: textbf
}% END: it
}% END: item
\\
Corrected, there is now a \emph{Markov models of codon evolution} subsection in the introductory chapter, which discusses most popular codon model implementations:

\begin{quote}
\myeditsvonea
\end{quote}

% 1b
\item {
{\it
\textbf{
Briefly described how to accommodate time-stamped sequences, which is now considered in the analyses in Chapter 2, but not really discussed.
}% END: textbf
}% END: it
}% END: item
\\
%
%PL: TODO: rewrite the final paragraph to reflect the dated-tip likelihood model: Essential, we now consider heights (instead of branch lengths) and introduce a new parameter, the rate of evolution r, which allows rescaling genetic distance into time. The task is now to estimate the heights and r from the sequence data, keeping the ti[ heights proportional to their sampling time, with the constant of proportionality being our estimable r.
%
Corrected, I added a subsection which describes the methods used in BEAST to accomodate time-stamped sequences as a source of calibration for trees, when such information is availiable:

\begin{quote}
\myeditsvoneb
\end{quote}

% 1c
\item {
{\it
\textbf{
Give a bit more detail on phylogeography, e.g. how do the models look like and what is the connection between phylogeography and epoch models (some connection is described in Chapter 2, but remains a bit vague).
}% END: textbf
}% END: it
}% END: item
\\
I added a figure and additional explanation, which should make the subsequent chapters clearer now. 
These changes start with:

\begin{quote}
\myeditsvonec
\end{quote}

### 2015-02-11 chapters/chapter01.tex

\subsection{Markov models of codon evolution\label{sub:codon}}
%TODO: briefly introduce, mg is later, show gy94

The first two computationally tractable codon substitution models were independently proposed by \cite{Muse1994} and \cite{Goldman1994} in the same issue of Molecular Biology and Evolution (MBE).
In these models, a non-stop codon triplet $n_{1}n_{2}n_{3}$ is considered to be the smallest unit of evolution.
According to the universal genetic code, there are $4^3$ possible triplets minus three stop codons, resulting in a state space size of $61$ codons.
The models inherit the standard assumption of independence, i.e. the substitutions at these three codon positions occur independently, and only a single change per triplet can occur at a given time. 
%%GB: explain in more detail, i.e. more than 1 change is possible in subsequent infinitesimal time intervals
Proteins are coded by a set of 20 amino acids, with each amino acid being coded by a codon triplet. 
Because there are 61 non-stop codons to encode for 20 amino acids, some amino acids will inevitably be coded by more than one codon.
A substitution that does not change the encoding amino-acid is called a \emph{synonymous} substitution, while a \emph{non-synonymous} substitution refers to a substitution that does result in an amino acid substitution.
Quantifying the rate at which these substitutions occur allows us to asses the nature of selective pressure, which is the main driving force behind molecular evolution.

@@

\subsection{Dated phylogenetic trees\label{sub:dated_trees}}
%how to accommodate time-stamped sequences

Time-stamped or `heterchronous' sequence data, which frequently arises from studies involving MEP populations for which sequences are sampled at different points in time and the evolutionary rates are high enough for significant changes to occur between the sampling times, can be used to calibrate the molecular clock and convert branch lengths into actual time units.

Fo simplicity let us assume a single site, where character sequence data was sampled on dates ranging from $1983$ to $2000$, as seen in Figure~\ref{fig:timing_sequences}.
We also assume that the rooted tree topology is known and we want to estimate the unknown times $T_{0},T_{6},T_{7},T_{8}$.

>>>>

\subsection{Markov models of codon evolution\label{sub:codon}}

\myedit{svonea}{
First two computationally tractable codon substitution models were independently proposed by \cite{Muse1994} and \cite{Goldman1994} in the same issue of Molecular Biology and Evolution (MBE).
In these models, a non-stop codon triplet $n_{1}n_{2}n_{3}$ is considered to be the smallest unit of evolution.
According to the universal genetic code, there are $4^3$ possible triplets minus three stop codons, resulting in a state space size of $61$ codons.
The models inherit the standard assumption of independence, i.e. the substitutions at these three codon positions occur independently, and only a single change per triplet can occur at a given time. 
%%GB: explain in more detail, i.e. more than 1 change is possible in subsequent infinitesimal time intervals
Proteins are coded by a set of 20 amino acids, with each amino acid being coded by a codon triplet. 
Because there are 61 non-stop codons to encode for 20 amino acids, some amino acids will inevitably be coded by more than one codon.
A substitution that does not change the encoding amino-acid is called a \emph{synonymous} substitution, while a \emph{non-synonymous} substitution refers to a substitution that does result in an amino acid substitution.
Quantifying the rate at which these substitutions occur allows us to asses the nature of selective pressure, which is the main driving force behind molecular evolution.
}

@@

\subsection{Dated phylogenetic trees\label{sub:dated_trees}}

\myedit{svoneb}{
Time-stamped or `heterchronous' sequence data, which frequently arises from studies involving MEP populations for which sequences are sampled at different points in time and the evolutionary rates are high enough for significant changes to occur between the sampling times, can be used to calibrate the molecular clock and convert branch lengths into actual time units.
}
For simplicity let us assume a single site, where character sequence data was sampled on dates ranging from $1983$ to $2000$, as seen in Figure~\ref{fig:timing_sequences}.
We also assume that the rooted tree topology is known and we want to estimate the unknown times $T_{0},T_{6},T_{7},T_{8}$.
